Supplementary Information: Drought sensitivity is not a species trait: Tropical tree drought sensitivity is jointly shaped by drought characteristics, species adaptations, and individual microenvironments

### 2025-09-17 doc/response_to_reviewer_R1.Rmd

---
title: 'Responses to the Reviewer'
output: 
    bookdown::word_document2: 
    #   reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
      reference_docx: word-styles-reference-singlespace.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
    #pdf_document:
    # engine: xelatex 
  #officedown::rdocx_document
bibliography: growth-precip-thailand.bib
# csl: new-phytologist.csl
csl: ecology-letters.csl 
suppress-bibliography: TRUE
---

@@

•	Since the study is based on only one site, it is relevant to improve the discussion section with other already published study sites (add more references), mainly link the global implications, such mortality rate and vulnerability in tropical forests. See https://urldefense.proofpoint.com/v2/url?u=https-3A__doi.org_10.1126_science.adq6607&d=DwIFaQ&c=009klHSCxuh5AI1vNQzSO0KGjl4nbi2Q0M1QLJX9BeE&r=6fH468T58at0TtM59sMf109A_dRgo6XTfGMOVX5Ewrw&m=fPE_4y1UPsTl9St_Y44IUO7VIaQJOJIUiHtK15-Nr7g4UiVKbtYQ-TE0043SLAj6&s=SxqVZgFkbqAZ0-VEvdQntXWpGEUu9lr25rgxQtwaVdw&e=

*RESPONSE:*

•	There was something missing to discuss about the species in particular, some that have been more affected than others and why, especially those in conservation categories.

**RESPONSE: Thanks for this suggestion. We have now included a supplementary table (Table S1) that details the response of specific species to specific drought events. This summary highlights our main message that the species responses are dissimilar across drought events. This table also includes available information on the IUCN categories of the species analysed (accessed in September 2025) where information was avaialable. Two of the species analysed are "Endangered", including *Afzelia xylocarpa*, the species that was discussed. There are also "Vulnerable" and "Data Deficient" species in the list, some of which have consistent negative drought sensitivities. This table would therefore be useful for forest managers and conservation professionals and to guide species-specific research directions. We have also added a line highlighting the contrasts between species responses in the discussion, which now reads:**

"An important implication is that drought responses inferred from tree-ring records derived from the few species known to form reliable, dateable annual rings (n=4 at HKK; @vlam_temperature_2014) are unlikely to be representative of the drought responses of the entire community.
For example, one of the few annual-ring-forming species at HKK, deciduous *Afzelia xylocarpa*, had the most distinct response across the three events; it's growth nearly doubled in the 2010 and 2020 droughts but was severely reduced in the wet-season drought (Fig. S5, Table S1).
Extrapolating from these could lead to erroneous inferences on species with conservation or commercial significance, e.g. *Dipterocarpus alatus* which presents a different drought response (Table S1)."

**As to why different species might be responding differently, a key difference could be their evolutionary strategies along hydraulic safety vs. efficiency. 
We have discussed our results in light of deciduousness as an example.
However, we recognise that many other above and belowground traits may contribute to these differences and need to examined further.
We have edited text in the discussion to read:**

"Paralleling the rarity of long-term records of annual tree growth in less seasonal tropical forests [@zuidema_tropical_2022], we know exceedingly little about how tropical tree species with different above- and below-ground traits along the hydraulic safety-efficiency spectrum respond to different types of drought, and how these responses may be modified by microenvironments."

@@

•	It is necessary to discuss further why the results indicate a "spatiotemporal insurance", which ecophysiological mechanisms or processes are relevant in these tropical forests.

**RESPONSE: Thanks for this suggestion. Our conclusion here was based on the fact that both species and individuals within species (spatial) seem to have little similarity in their responses to different droughts. At the community/stand level, these could be either because of adaptive differences among species or environmental differences among sites that interact with different drought types. Upon further discussion, we think that the term "biological insurance" (*sensu* @loreau_biodiversity_2021) may be more appropriate here given its more general definition and inclusion of both species and spatial concepts of insurance. We have now revised the usage of this term in the manuscript and included a reference to this framework.**

@@

2 The topographic wetness index. Are there any measured soil moisture data in this plot? Since this index is only based on topography, it will be good if the author could provide some measured soil moisture data to calibrate the prediction based on model prediction.

**RESPONSE: Unfortunately soil moisture data from the plot is not available for the years of study across which dendroband data are available. HKK, being fairly remote, has limited access to power sources and other infrastructure that are needed to maintain such data collection. We are currently piloting a few basic soil moisture sensors, however, this will have limited value for calibration as they only measure the top layer of the soil and are limited in spatial scale compared to the spatial resolution of TWI data.**

3 In the case of deciduousness, the authors indicated that they used the deciduous index from Williams et al. (2008). Since different species might shed leaves in different season. Did the authors consider the impact of the timing of deciduousness on the growth?

*RESPONSE: Thanks for this interesting comment. We agree that there are several hydraulic strategies both above and belowground that might impact tree growth sensitivity to drought. We chose degree of deciduousness as a simple, illustrative example of a strategy that might be associated with some of this variation. We agree that the timing of deciduousness may be an important factor to consider, as might leaf turgor loss point, wood density etc. and considering these would be a key direction for future analyses, but beyond the scope of the current analysis.*

### 2025-09-17 doc/manuscript_hkk_drought_sensitivity.Rmd
update with edit to species response comment

<!----Limitations in generalisation/ species examined here may have greater diversity of drought adaptaitons than elsewhere in the tropics---->
Periodic hot/dry conditions in tropical dry forests could create selective pressures for adaptative variation towards drought resistance potentially resulting in a greater diversity of drought responses compared to other bioclimatic zones where adaptive pressures and thus ecological variability are different. 
We demonstrate that species identity and a key species adaptation (decidousness) interact with drought characteristics and microenvironment to shape drought responses in a dry seasonal evergreen forest.
This aligns closely with recent global analyses using tree ring chronologies showing hetereogeneous responses of tree functional types to drought types [@zuidema_pantropical_2025].
Strong seasonality in tropical dry forests exerts selective pressures on hydraulic strategies resulting in diversification along the hydraulic safety-efficiency axis with diverse strategies to optimise tree vital rates [@gonzález-m._diverging_2021;@oliveira_linking_2021]. 
The degree of adaptive variation along the water-use axis might be comparable to other dry climatic systems (e.g. temperate drylands reported in @vasey_intraspecific_2022), but may be distinct from the wet/everwet tropics, where tree-environment relations and evolutionary strategies may be substantially different. 
Paralleling the rarity of long-term records of annual tree growth in less seasonal tropical forests [@zuidema_tropical_2022], we know exceedingly little about how tropical tree species with different traits respond to different types of drought, and how these responses may be modified by microenvironments.
Again, continued expansion of dendrometer band records and ecophysiological studies will be essential to understanding the diversity and complexity of tropical tree drought responses, as well as if and how this diversity may promote community-level insurance effects under drought across tropical forests in general.
<!--Therefore, we caution that our results suggestive of community-level insurance effects under drought be interpreted carefully across forest types that may be structured primarily by nutrient limitation, fires, herbivory or pathogens. 
However, our inference on the effects of drought timing could be generalisable across wet and dry tropics in the light of emerging evidence that despite low seasonality in climatic conditions, everwet forest species also show seasonal growth patterns [@giraldo_tree_2023].   -->

@@

<!----Contributions beyond---->
Despite consistent observations of plot-level growth declines across drought events, we show that this scale obscures a variety of species- and tree-level ecological responses that are heterogeneous across droughts. 
An important implication is that drought responses inferred from tree-ring records derived from the few species known to form reliable, dateable annual rings (n=4 at HKK; @vlam_temperature_2014) are unlikely to be representative of the drought responses of the entire community.
For example, one of the few annual-ring-forming species at HKK, deciduous *Afzelia xylocarpa*, had the most distinct response across the three events; it's growth nearly doubled in the 2010 and 2020 droughts but was severely reduced in the wet-season drought (Fig. S5, Table S1).
More generally, tree-ring records likely present a biased picture of drought responses, as ring formation is caused by dry-season dormancy (often associated with deciduous leaf habit) and sampling often targets large trees with exposed crowns and dry microenvironments [@speer_fundamentals_2010]-- all factors shown here to <!--directly and interactively -->influence drought sensitivity.
<!-- More generally, past analyses of tree ring and census estimates of drought sensitivity in seasonally dry tropical forests may have been limited towards fast growing (measurable using tape) or few dry-associated species (that form annual tree rings) - and therefore capturing only the more dramatic effects. -->
Our analyses reveal a complex heterogeneity of drought responses, including the contrasting influence of deciduousness in three different droughts, showing that ecological and evolutionary strategies can make particular species and trees resistant to some droughts but vulnerable to others. 
These distinct responses of species and individuals to droughts suggests potential biological insurance effects among tropical tree responses, in alignment with recent global analyses [@liu_drought_2022; @zuidema_pantropical_2025].

>>>>

<!----Limitations in generalisation/ species examined here may have greater diversity of drought adaptaitons than elsewhere in the tropics---->
Periodic hot/dry conditions in tropical dry forests could create selective pressures for adaptative variation towards drought resistance potentially resulting in a greater diversity of drought responses compared to other bioclimatic zones where adaptive pressures and thus ecological variability are different. 
We demonstrate that species identity and a key species adaptation (decidousness) interact with drought characteristics and microenvironment to shape drought responses in a dry seasonal evergreen forest.
This aligns closely with recent global analyses using tree ring chronologies showing hetereogeneous responses of tree functional types to drought types [@zuidema_pantropical_2025].
Strong seasonality in tropical dry forests exerts selective pressures on hydraulic strategies resulting in diversification along the hydraulic safety-efficiency axis with diverse strategies to optimise tree vital rates [@gonzález-m._diverging_2021;@oliveira_linking_2021]. 
The degree of adaptive variation along the water-use axis might be comparable to other dry climatic systems (e.g. temperate drylands reported in @vasey_intraspecific_2022), but may be distinct from the wet/everwet tropics, where tree-environment relations and evolutionary strategies may be substantially different. 
Paralleling the rarity of long-term records of annual tree growth in less seasonal tropical forests [@zuidema_tropical_2022], we know exceedingly little about how tropical tree species with different above- and below-ground traits along the hydraulic safety-efficiency spectrum respond to different types of drought, and how these responses may be modified by microenvironments.
Again, continued expansion of dendrometer band records and ecophysiological studies will be essential to understanding the diversity and complexity of tropical tree drought responses, as well as if and how this diversity may promote community-level insurance effects under drought across tropical forests in general.
<!--Therefore, we caution that our results suggestive of community-level insurance effects under drought be interpreted carefully across forest types that may be structured primarily by nutrient limitation, fires, herbivory or pathogens. 
However, our inference on the effects of drought timing could be generalisable across wet and dry tropics in the light of emerging evidence that despite low seasonality in climatic conditions, everwet forest species also show seasonal growth patterns [@giraldo_tree_2023].   -->

@@

<!----Contributions beyond---->
Despite consistent observations of plot-level growth declines across drought events, we show that this scale obscures a variety of species- and tree-level ecological responses that are heterogeneous across droughts. 
An important implication is that drought responses inferred from tree-ring records derived from the few species known to form reliable, dateable annual rings (n=4 at HKK; @vlam_temperature_2014) are unlikely to be representative of the drought responses of the entire community.
For example, one of the few annual-ring-forming species at HKK, deciduous *Afzelia xylocarpa*, had the most distinct response across the three events; it's growth nearly doubled in the 2010 and 2020 droughts but was severely reduced in the wet-season drought (Fig. S5, Table S1).
Extrapolating from these could lead to erroneous inferences on species with conservation or commercial significance, e.g. *Dipterocarpus alatus* which presents a different drought response (Table S1).
More generally, tree-ring records likely present a biased picture of drought responses, as ring formation is caused by dry-season dormancy (often associated with deciduous leaf habit) and sampling often targets large trees with exposed crowns and dry microenvironments [@speer_fundamentals_2010]-- all factors shown here to <!--directly and interactively -->influence drought sensitivity.
<!-- More generally, past analyses of tree ring and census estimates of drought sensitivity in seasonally dry tropical forests may have been limited towards fast growing (measurable using tape) or few dry-associated species (that form annual tree rings) - and therefore capturing only the more dramatic effects. -->
Our analyses reveal a complex heterogeneity of drought responses, including the contrasting influence of deciduousness in three different droughts, showing that ecological and evolutionary strategies can make particular species and trees resistant to some droughts but vulnerable to others. 
These distinct responses of species and individuals to droughts suggests potential biological insurance effects [@loreau_biodiversity_2021] among tropical tree responses, in alignment with recent global analyses [@liu_drought_2022; @zuidema_pantropical_2025].
